---
title: "Building a GRanges Reference Object"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```

::: callout-note
Code in this chapter is shown but not executed. The steps depend on large reference files and cluster resources that are not available at render time. Paths are written generically; substitute your own.
:::

## Building a Reference Object

Downstream ATAC processing through `Signac` requires specific annotations in the form of a `GRanges` object carrying gene coordinates, IDs, and a `gene_biotype` column, which drives gene activity scoring and coverage plotting. The Lawson Lab annotations use in-house Lawson nomenclature, which is why 3 separate GTF files are required: one which carries the coordinates and features (shared with the features in your counts matrix), one which serves as a map between Lawson nomenclature and Ensembl nomenclature, and one native GRCz11 gtf which contains crucial biotype information. We use these 3 files together to collate all the necessary information per gene to build a master GRanges object.

### What you need

1.  `lawson_zebrafish.gtf`: Contains the coordinates and features (i.e., the gene IDs appearing in the counts matrix).
2.  `V4.3.2.validated.gtf`: Another Lawson Lab file containing *both* Lawson gene IDs and Ensembl transcript IDs in the same row. This file serves as a dictionary for mapping between Lawson nomenclature and Ensembl nomenclature.
3.  `Danio_rerio.GRCz11.113.gtf`: This file carries gene biotype information, which will be required downstream for ATAC pre-processing functionality in `Signac`.

::: callout-warning
File 1 must be the same GTF that was passed to `cellranger-arc mkref`. If it is not, the gene IDs in the annotation object will not match the features in the counts matrix.
:::

### Software

::: callout-important
TODO: pin a reproducible environment. Options are a `renv.lock` committed alongside the book, a Conda environment YAML, or a container definition. Whichever is chosen, record the Bioconductor release as well as the package versions, since `rtracklayer` and `GenomicRanges` are versioned together.
:::

```{r}
#| label: libraries
library(rtracklayer)
library(GenomicRanges)
```

### Setting paths

```{r}
#| label: annotation-paths
refDir <- "/users/mike/Desktop/multiome-tutorial/references/"

# Coordinates and features; the GTF used to build the CellRanger reference
lawsonGTFpath <- paste0(refDir, "GTFs/lawson_zebrafish.gtf")

# Dictionary: Lawson gene_id <-> Ensembl transcript_id
llV4path <- paste0(refDir, "GTFs/V4.3.2.validated.gtf")

# Reference annotation carrying gene_biotype
GRCz11path <- paste0(refDir, "GTFs/Danio_rerio.GRCz11.113.gtf")
```

### Building the biotype dictionary

The chain runs in three steps. Ensembl supplies the biotype for each transcript ID. The validated Lawson GTF maps those transcript IDs to Lawson gene IDs. The result is a lookup table keyed on Lawson gene IDs.

```{r}
#| label: biotype-lookup
# Ensembl: transcript_id -> gene_biotype
ens   <- rtracklayer::import(GRCz11path, format = "gtf")
ensTx <- as.data.frame(mcols(ens))
ensTx <- ensTx[ensTx$type == "transcript", c("transcript_id", "gene_biotype")]

# Lawson validated: gene_id <-> transcript_id
ll   <- rtracklayer::import(llV4path, format = "gtf")
llTx <- as.data.frame(mcols(ll))
llTx <- unique(llTx[llTx$type == "transcript", c("gene_id", "transcript_id")])

# Join and collapse to one biotype per Lawson gene_id
map <- merge(llTx, ensTx, by = "transcript_id", all.x = TRUE)
map <- map[!is.na(map$gene_biotype), c("gene_id", "gene_biotype")]
map <- map[!duplicated(map$gene_id), ]
```

The collapse on the last line is a decision rather than bookkeeping. A gene with several transcripts can map to more than one biotype. Taking the first asserts that any gene with a protein-coding transcript is a protein-coding gene. That is usually the intended behavior, but it is worth knowing how often it actually applies:

```{r}
#| label: biotype-conflicts
conflicts <- merge(llTx, ensTx, by = "transcript_id", all.x = TRUE)
conflicts <- conflicts[!is.na(conflicts$gene_biotype), ]
multi <- tapply(conflicts$gene_biotype, conflicts$gene_id,
                function(x) length(unique(x)))
table(multi)
```

### Applying biotypes to the Lawson annotation

```{r}
#| label: apply-biotypes
annotations <- rtracklayer::import(lawsonGTFpath)

idx <- match(mcols(annotations)$gene_id, map$gene_id)
mcols(annotations)$gene_biotype <- map$gene_biotype[idx]
mcols(annotations)$gene_biotype[is.na(mcols(annotations)$gene_biotype)] <- "Other"
```

Custom project-specific features (transgenes, reporters, knock-in constructs) have no Ensembl counterpart, so they fall through to `"Other"` and must be set explicitly. Match on `gene_id`, never on row position.

```{r}
#| label: custom-features
transgenes <- c("<transgene_1>", "<transgene_2>")

stopifnot(all(transgenes %in% mcols(annotations)$gene_id))

mcols(annotations)$gene_biotype[
  mcols(annotations)$gene_id %in% transgenes] <- "protein_coding"
```

::: callout-warning
Do not assign biotypes by row index. Row positions shift whenever the GTF is regenerated, a release is bumped, or a join expands and collapses the table. Indexing by position silently relabels arbitrary genes elsewhere in the genome, with no error to catch it. The `stopifnot()` above exists so that a renamed or missing transgene fails loudly instead.
:::

### Finalizing the object

```{r}
#| label: finalize-annotation
keep <- c("type", "gene_id", "gene_name", "transcript_id", "gene_biotype")
mcols(annotations) <- mcols(annotations)[, keep]

genome(annotations) <- "GRCz11"

saveRDS(annotations,
        file = file.path(refDir, "Lawson_annotation_with_biotypes.Rds"))
```

Two details cause trouble later if skipped. Setting `genome()` matters because `Annotation()<-` and `TSSEnrichment()` warn or misbehave without it. And the sequence naming style has to match the fragments file:

```{r}
#| label: seqlevels-check
seqlevelsStyle(annotations)
head(seqlevels(annotations))
```

Ensembl-derived annotations use `1`, `2`, `3`; UCSC-derived files use `chr1`, `chr2`, `chr3`. A mismatch between the annotation and the fragments file produces zero overlaps rather than an error, which surfaces much later as an empty gene activity matrix or a TSS enrichment score of zero. Convert before saving if they differ:

```{r}
#| label: seqlevels-fix
seqlevelsStyle(annotations) <- "Ensembl"
```

### Checks before moving on

```{r}
#| label: annotation-checks
table(mcols(annotations)$gene_biotype)
mcols(annotations)[mcols(annotations)$gene_id %in% transgenes, ]
length(unique(mcols(annotations)$gene_id))
```

Confirm that the biotype table is not overwhelmingly `"Other"`, that the custom features are labeled as intended, and that the gene count falls in the expected range for the annotation. A large `"Other"` fraction usually means the dictionary join failed, most often because the two GTFs carry different identifier versions (`ENSDART00000000001` against `ENSDART00000000001.1`). Strip the version suffix and rejoin if so.

## RNA Preprocessing

## ATAC Preprocessing
