---
title: "Building a GRanges Reference Object"
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(eval = FALSE)
```

::: callout-note
Code in this chapter is shown but not executed. The steps depend on large reference files and cluster resources that are not available at render time. Paths are written generically; substitute your own.
:::

## Building a Reference Object

Downstream ATAC processing through `Signac` requires specific annotations in the form of a `GRanges` object carrying gene coordinates, IDs, and a `gene_biotype` column. The Lawson Lab annotations use in-house nomenclature, which is why 3 separate GTF files are used: one which carries the coordinates and features (what you used in the CellRanger ARC pipeline), one which serves as a map between Lawson nomenclature and Ensembl nomenclature, and one native GRCz11 gtf which contains crucial biotype information. We use these 3 files together to collate all the necessary information per gene to build a master `GRanges` object.

### What is a GRanges object and why do we need one?

A `GRanges` object is Bioconductor's container for genomic intervals. Each element in a `GRanges` object contains at minimum a sequence name, the start and end coordinates, the strand, and abritrary metadata columns. The reason this matters more for the ATAC modality is based in the nature of the assay. Gene expression is a relatively simple gene-by-cell matrix and values represent counts for that gene in that cell. ATAC data instead outputs fragments which are essentially millions of genomic intervals with no feature label attached. In order to assign our peaks to genomic features such as genes and promoters, we need to provide annotations that inform *where* genes are. `Signac` (the package used for the ATAC modality), internally uses `GRanges` for various downstream pre-processing functions and as such, this is a crucial and non-negotiable step.

::: callout-note
## The worked example

The Cell Ranger run this chapter follows was annotated against the reference at:

```         
/dartfs-hpc/rc/lab/G/GSR_Active/scripts/refs/Danio_rerio/Lawson/GRCz11_lawson
```

which was built from:

```         
/dartfs-hpc/rc/lab/G/GSR_Active/scripts/refs/Danio_rerio/Lawson/5b_V4.3.2finalFixChrOnly.gtf
```

renamed to `lawson_zebrafish.gtf` for readability. Two things are worth reading off that filename. It is a V4.3.2 derivative, so it shares lineage with the dictionary file below. And `FixChrOnly` means it is restricted to assembled chromosomes and uses `chr`-prefixed sequence names, which matters in the final step.
:::

### What you need

1.  `lawson_zebrafish.gtf`: This is the file you used to run `CellRanger mkref`. This contains the coordinates and features (i.e., the gene IDs appearing in the counts matrix), along with any custom gene annotations.
2.  `V4.3.2.validated.gtf`: Another Lawson Lab file containing *both* Lawson gene IDs and Ensembl transcript IDs in the same row. This file serves as a dictionary for mapping between Lawson nomenclature and Ensembl nomenclature.
3.  `Danio_rerio.GRCz11.113.gtf`: This file carries gene biotype information, which will be required downstream for ATAC pre-processing functionality in `Signac`.

::: callout-warning
File 1 must be the same GTF that was passed to `cellranger-arc mkref`. If it is not, the gene IDs in the annotation object will not match the features in the counts matrix.
:::

### Do you actually need all three?

Often not. We need to have **coordinates** (meaning we need a GTF whose gene IDs match your counts matrix), and **biotypes**, which only Ensembl supplies. The dictionary file is needed solely when the coordinate GTF does not already carry Ensembl transcript IDs.

The three-file version below is written against the standalone Lawson GTF because that is the more general situation, and because it makes the mapping between nomenclatures explicit. In this example, we have added 2 custom transgenes to the gtf which need to be accounted for.

### Software

::: callout-important
TODO: pin a reproducible environment. Options are a `renv.lock` committed alongside the book, a Conda environment YAML, or a container definition. Whichever is chosen, record the Bioconductor release as well as the package versions, since `rtracklayer` and `GenomicRanges` are versioned together.
:::

```{r}
#| label: libraries
library(tidyverse)
library(GenomicRanges)
library(rtracklayer)

set.seed(1234)
```

### Setting paths

```{r}
#| label: annotation-paths
#----- Set paths
wd <- "/Volumes/GMBSR_bioinfo/Labs/kasper/250613_10x_multiome_redo/"
files <- paste0(wd, "files/")
filesDir <- paste0(files, "lawson_db_info/")

#----- Create directories
create <- c(files, filesDir, paste0(filesDir, "Database_Info/"))
for (i in create) {
  if (!dir.exists(i)) {
    dir.create(i)
  }
}
```

```{r}
#| label: gtf-paths
lawsonGTFpath <- paste0(filesDir, "GTFs/lawson_zebrafish.gtf") # Lawson
llV4path <- paste0(filesDir, "GTFs/V4.3.2.validated.gtf") # Lawson-Ensembl TXIDs
GRCz11path <- paste0(filesDir, "GTFs/Danio_rerio.GRCz11.113.gtf") # Ensembl
```

### Reading in the GTF files as GRanges Objects

`rtracklayer::import()` parses the GTF into a `GRanges` object. The metadata columns are pulled out into a plain data frame so they can be joined against the other two annotations. We can read in all three GTFs in the same way.

```{r}
#| label: process-lawson
#----- Read in the lawson annotations
annotations <- rtracklayer::import(lawsonGTFpath)

#----- Extract metadata
annoData <- as.data.frame(mcols(annotations))
```

Now, we read in the second Lawson Lab file which contains the mapping info. The only difference here is that we subset to just transcript rows since we are interested in linking gene IDs to Ensembl transcript IDs. By doing this, every transcript with at least one exon is represented.

```{r}
#| label: process-txid-map
#----- This reference will map all Lawson annotations to Ensembl TXID counterparts
ll <- rtracklayer::import(llV4path, format = "gtf")
llmeta <- as.data.frame(mcols(ll)) %>%
  distinct()
llmeta <- llmeta[llmeta$type == "transcript",]
```

Next, we read in the GRCz11 Ensembl annotations the same way.

```{r}
#| label: process-ensembl
#----- Read in the Zebrafish reference gtf from Ensembl
grcz11gtf <- rtracklayer::import(GRCz11path, format = "gtf")
```

### Link metadata

Ensembl supplies the protein-coding transcript IDs. Those are matched into the Lawson-Ensembl map to find which Lawson gene IDs they belong to, and those gene IDs are labeled `protein_coding`. Everything else becomes `Other`.

```{r}
#| label: link-metadata
#----- Extract gene biotype and ID information
ensemblMetadata <- as.data.frame(mcols(grcz11gtf)) %>%
  distinct()
ensMeta <- ensemblMetadata[ensemblMetadata$type == "transcript",]

#----- Get just the protein coding genes
pcode <- ensMeta[ensMeta$gene_biotype == "protein_coding",]
txs <- pcode$transcript_id

#----- Filter llmeta to just include those transcripts
llsubset <- llmeta[llmeta$transcript_id %in% txs,]
llsubset$gene_biotype <- "protein_coding"

#----- Merge
merged <- annoData %>%
  left_join(llsubset[,c("gene_id", "gene_biotype")], by = "gene_id", relationship = "many-to-many")
merged$gene_biotype.x <- NULL

#----- Rename the column to gene_biotype
colnames(merged)[6] <- c("gene_biotype")

#----- Set NA to other
merged$gene_biotype[is.na(merged$gene_biotype)] <- "Other"
fileName <- paste0(filesDir, "Protein_Coding_Genes.csv")
write.csv(merged, fileName)
```

### Reconstruct GRanges object

The coordinates are read back from the GTF with `read.table()` and bound to the metadata, then the biotype column is joined on and the whole thing is rebuilt as a `GRanges`.

```{r}
#| label: reconstruct-granges
#----- Read in the original gtf file
gtf <- read.table(lawsonGTFpath, sep = "\t")

#----- Extract GRanges information
gtf <- gtf[,1:8]

#----- Combine the Granges information and the original metadata
full <- cbind(gtf, annoData)

#----- Merge
final <- full %>%
  left_join(merged[,c("gene_id", "gene_biotype")], by = "gene_id", relationship = "many-to-many") %>%
  distinct()
final$gene_biotype.x <- NULL
colnames(final)[14] <- c("gene_biotype")

#----- Change the custom annotations for the project to protein coding
final$gene_biotype[final$gene_id %in% c("EGFP", "Kalta4")] <- c("protein_coding")

#----- Save as a csv
dbFile <- paste0(filesDir, "Database_Info/Lawson_Reference_GRanges_Meta_with_GeneBiotypes.csv")
write.csv(final, file = dbFile)

#----- Convert back to a GRanges object
gr <- GRanges(
  seqnames = final$V1,
  ranges = IRanges(start = final$V4, end = final$V5),
  strand = final$V7,
  gene_id = final$gene_id,
  gene_name = final$gene_name,
  transcript_id = final$transcript_id,
  gene_biotype = final$gene_biotype,
  type = final$type
)

#----- Save granges object as an rds
rdsFile <- paste0(filesDir, "Lawson_Reference_GenomicRanges_with_GeneBiotype.Rds")
saveRDS(gr, file = rdsFile)
```

``` bash
```
