---
title: "03-RNA-pre-procesing-and-QC"
---

```{r}
#| label: setup 
#| include: false 
knitr::opts_chunk$set(eval = FALSE)
```

::: callout-note
Code in this chapter is shown but not executed. The steps depend on large reference files and cluster resources that are not available at render time. Paths are written generically; substitute your own.
:::

## Software


## What files do we need?

Gene expression counts live in the CellRanger output folder under the `outs` folder as `filtered_feature_bc_matrix.h5`. This is an HDF5 container that contains a sparse matrix and additional row/column information. Here, columns are cells and rows are features (genes). A sparse matrix means that only non-zero entries are stored which greatly reduces file size. 

CellRanger ARC internally calls cells by leveraging information from both modalities to determine which barcodes are valid nuclei. This is what the `filtered` portion of the file name implies. The counts represented within this file are UMI-deduplicated, confidently mapped, and collapsed to the gene level (for the GEX portion) against the GTF used. The alternative file, `raw_feature_bc_matrix.h5` instead contains every barcode which includes empty droplets carrying ambient RNA. This is useful for assessing a potential ambient RNA profile in your data, but it not a good starting point for gene expression analysis. 

Typically, a regular single-cell RNAseq experiment `filtered_feature_bc_matrix.h5`, only carries gene expression information, however for a Multiome experiment, it carries both the GEX and ATAC data. As such, we need to specify when reading in the files that we are interested in pulling the list element corresponding to gene expression. 

When we have multiple samples present, it is often helpful to define utility functions which we can use to map across our samples. This not only saves us time, but keeps our code files clean by avoiding redundant lines. 

### Reading in filtered feature barcode matrices

First, let's define a function to read in the `filtered_feature_bc_matrix.h5` files.
This function takes 2 arguments:

1. `sampleDir`: a string ending in "/" pointing to the CellRanger output folder

2. `sample`: The name of the individual sample (i.e., a subdirectory of `sampleDir`)

This function constructs the path to that sample's filtered barcode matrix and creates a `Seurat` object using it's `Gene Expression` element. Here, we hardode that we want to preliminarily filter the sample for `min.cells = 10` and `min.features = 100` (the former indicates that a feature must be present in at least 10 cells to be retained and the latter indicates that a cell must contain at least 100 features to be retained). By doing this, we preliminarily remove cells with very low genomic complexity. 

The function also calculates the percentage of mitochondrial features using zebrafish-specific mitochondrial prefixing (lower case "mt") as well as other complexity metrics such as `log10GenesPerUMI` which is a complexity measure. 


```{r}
#| label: 10x reading utility function

#----- Function to read in filtered feature barcode matrices
read_10x_multiGEX <- function(sampleDir, sample) {
  message(paste0("Reading sample: ", sample))
  
  
  #----- CREATE SEURAT OBJECT
  counts <- Read10X_h5(filename = paste0(sampleDir, sample, "/outs/filtered_feature_bc_matrix.h5"))
  x <- CreateSeuratObject(counts = counts$`Gene Expression`,
                          min.cells = 10,
                          min.features = 100,
                          assay = "RNA")
  x$Ident <- sample
  #----- ADD PERCENT MITOCHONDRIAL GENES
  x$percentMT <- PercentageFeatureSet(object = x,
                                         pattern = "^mt-")
  x$log_nCount_RNA <- log(x$nCount_RNA)
  x$log_nFeature_RNA <- log(x$nFeature_RNA)
  #----- ADD LIBRARY COMPLEXITY METRIC
  x$log10GenesPerUMI <- log10(x$nFeature_RNA) / log10(x$nCount_RNA)
  print(ncol(x))
  return(x)
}

```

Given a vector of sample names (corresponding to how our CellRanger output folder subdirectories are named), we can map this function using `purrr` which will return a list of `Seurat` objects.

```{r}
#| label: Read-files

#----- Set working directory and sample directory
wd <- "/Volumes/GMBSR_bioinfo/Labs/kasper/250613_10x_multiome_redo/"
sampleDir <- paste0(wd, "raw_data/")

#----- Vector of samples
samples <- list("12hpf", "14hpf", "18hpf", "22hpf", "26hpf", "30hpf")
names(samples) <- c("1", "2", "3", "4", "5", "6")

#----- Read in files
seurat_list <- map(samples, ~ read_10x_multiGEX(sampleDir, .x))
names(seurat_list) <- samples

#----- Get combined metadata across all samples
metadata <- data.frame()
for (i in 1:length(seurat_list)) {
  seurat_list[[i]]$orig.ident <- idents[[i]]
  metadata <- rbind(metadata, seurat_list[[i]]@meta.data)
}

```

